Anomaly detection with kernel density estimation on manifolds

### 2022-06-07 paper/paper.Rmd
Change notation for \hat{f}

Consider a compact Riemannian manifold $(M, g)$ of dimension $d$ without boundary and a probability distribution with density $f$ on the manifold. Assume $(M, d_g)$ is a complete metric space, where $d_g$ is the Riemannian distance induced by the Riemannian, $g$, and the strictly positive injectivity radius [@Chavel2006-mp] of the manifold, $inj_g(M)$. Denote $A_i, i = 1, 2, \dots, N$, where $\pmb{A}_i \in \mathcal{R}^d$, as i.i.d. random objects on $M$ with density $f$. For each $p\in M$, @Pelletier2005-vu proposed the kernel density estimator of $f$ to be 
\begin{equation}
\label{eq:denriem}
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{A_i}(p)} K(\frac{d_g(p, A_i)}{h}),
\end{equation} 
where $h$ is the bandwidth satisfying $h \leq h_0$ for fixed $h_0$ such that $0\leq h_0\leq inj_g(M)$, $K: \mathcal{R}_{+} \rightarrow \mathcal{R}$ is a non-negative map, and $\theta_{q}(p)$ is the volume density function on the manifold.
@Pelletier2005-vu mention that the volume density function $\theta_q(p)$ defined for $p$ in a neighborhood of $q$ on the manifold, with the geodesic normal coordinates at $q$, is equal to the determinant of the Riemannian metric $g$ at the logarithm map at q $\exp_q^{-1}(p)$. The expression in \@ref(eq:denriem) is also proved to be consistent with the kernel density estimators in the Euclidean case as 
\begin{equation}
\label{eq:denconsist}
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d} K(\frac{\|p-A_i\|}{h}),
\end{equation} 
and converges at the same rate as the Euclidean kernel density estimator [@Henry2009-ll].

However, in application, we usually found the observations lie on a $d$-dimensional manifold embedded in a $s$-dimensional space, where $d \ll s$. Then manifold learning can be applies to find the $d$-dimensional representation of the manifold $M$. Therefore, based on the estimator in \@ref(eq:denriem), we propose to a kernel density estimator of f with the manifold embedding $y_i, i = 1, 2, \dots, N$ to be 
\begin{equation}
\label{eq:denestimator}
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h_i^d |g(\exp_p^{-1}(y_i))|} K(\frac{d_g(p, y_i)}{h_i}),
\end{equation} 
where $h_i, i = 1, 2, \dots, N$ is the pointwise variable bandwidth matrix and $h_i$ is positive definite with dimension
$d\times d$.

@@

## Riemannian metric as variable bandwidth

>>>>

Consider a compact Riemannian manifold $(M, g)$ of dimension $d$ without boundary and a probability distribution with density $f$ on the manifold. Assume $(M, d_g)$ is a complete metric space, where $d_g$ is the Riemannian distance induced by the Riemannian, $g$, and the strictly positive injectivity radius [@Chavel2006-mp] of the manifold, $inj_g(M)$. Denote $\pmb{X}_i, i = 1, 2, \dots, N$, where $\pmb{X}_i \in \mathcal{R}^d$, as i.i.d. random objects on $M$ with density $f$. For each point $p\in M$, @Pelletier2005-vu proposed a kernel density estimator of $f$ to be 
\begin{equation}
\label{eq:denriem}
\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{\pmb{X}_i}(p)} K(\frac{d_g(p, \pmb{X}_i}{h}),
\end{equation} 
where $h$ is the global bandwidth satisfying $h \leq h_0$ for fixed $h_0$ such that $0\leq h_0\leq inj_g(M)$, $K: \mathcal{R}_{+} \rightarrow \mathcal{R}$ is a non-negative map, and $\theta_{q}(p)$ is the volume density function on the manifold.
@Pelletier2005-vu mention that the volume density function $\theta_q(p)$ defined for $p$ in a neighborhood of $q$ on the manifold, with the geodesic normal coordinates at $q$, is equal to the determinant of the Riemannian metric $g$ at the logarithm map at q, $\exp_q^{-1}(p)$. The expression in \@ref(eq:denriem) is also proved to be consistent with the kernel density estimators in the Euclidean case as 
\begin{equation}
\label{eq:denconsist}
\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d} K(\frac{\|p-\pmb{X}_i\|}{h}),
\end{equation} 
and converges at the same rate as the Euclidean kernel density estimator [@Henry2009-ll].

### Volume density function

However, in application, we usually found the observations lie on a $d$-dimensional manifold embedded in a $s$-dimensional space, where $d \ll s$. Then manifold learning can be applies to find the $d$-dimensional representation of the manifold $M$. Therefore, based on the estimator in \@ref(eq:denriem), we propose a kernel density estimator of f with the manifold embedding $y_i, i = 1, 2, \dots, N$ to be 
\begin{equation}
\label{eq:denestimator}
\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h_i^d |g(\exp_p^{-1}(y_i))|} K(\frac{d_g(p, y_i)}{h_i}),
\end{equation} 
where $h_i, i = 1, 2, \dots, N$ is the pointwise variable bandwidth matrix and $h_i$ is positive definite with dimension $d\times d$.

@@

### Riemannian metric as variable bandwidth
